---
title: "Lakisha"
format: html
editor: visual
---

# Question 1

1.  Read the introduction and conclusion of BM. Then write a short paragraph answering the following:

    a.  What research question do BM try to answer?

    b.  What data and methodology do they use to address the question?

    c.  What do the authors consider to be their key findings?

# Question 2

```{r}
library(tidyverse)
bm <- read_csv('https://ditraglia.com/data/lakisha_aer.csv')
```

1.  Now that you have a rough idea of what the paper is about, it’s time to examine the dataset `bm`. Carry out the following steps:

    a.  Display the tibble `bm`. How many rows and columns does it have?

    b.  Display only the columns `sex`, `race` and `firstname` of `bm`. What information do these columns contain? How are `sex` and `race` encoded?

    c.  Add two new columns to `bm`: `female` should take the value `TRUE` if `sex` is female, and `black` should take value `TRUE` if `race` is black.

## Question 2a

```{r}
bm
```

4870 $\times$ 65

## Question 1b

```{r}
bm |>
  select(sex, race, firstname)

typeof(bm[["sex"]])
typeof(bm[["race"]])
typeof(bm[["firstname"]])
```

Sex (char): f - female, m - male

Race (char): w - white, b - black

firstname (char): first name of the candidate

## Question 1c

```{r}
bm <- 
  bm |>
    mutate(female = (sex == "f"),
         black = (race == "b"))
```

# Question 3

# Question 4

Randomized controlled trials are all about *balance*: when the treatment is randomly assigned, the characteristics of the treatment and control groups will be the same on average. To answer the following parts you’ll need a few additional pieces of information. First, the variable `computerskills` takes on the value `1` if a given resume says that the applicant has computer skills. Second, the variables `education` and `yearsexp` indicate level of education and years experience, while `ofjobs` indicates the number of previous jobs listed on the resume. To analyse whether characteristics are balanced across treatment and control, compare summary statistics between applicants with white- and black-sounding names.

a.  Is sex balanced across race? Use `dplyr` to answer this question. **Hint:** what happens if you apply the function `sum` to a vector of `TRUE` and `FALSE` values?

b.  Are computer skills balanced across race? **Hint:** the summary statistic you’ll want to use is the *proportion* of individuals in each group with computer skills. If you have a vector of ones and zeros, there is a very easy way to compute this.

c.  Are `education` and `ofjobs` balanced across race?

d.  Compute the mean and standard deviation of `yearsexp` by race. Comment on your findings.

e.  Why do we care if `sex`, `education`, `ofjobs`, `computerskills`, and `yearsexp` are balanced across race?

f.  Is `computerskills` balanced across `sex`? What about `education`? What’s going on here? Is it a problem? **Hint:** re-read section II C of the paper.
