---
title: "Lakisha"
author: "Xiaotian Tian"
format: pdf
editor: visual
---

# Question 1

1.  Read the introduction and conclusion of BM. Then write a short paragraph answering the following:

    a.  What research question do BM try to answer?

    b.  What data and methodology do they use to address the question?

    c.  What do the authors consider to be their key findings?

# Question 2

```{r}
library(tidyverse)
bm <- read_csv('https://ditraglia.com/data/lakisha_aer.csv')
```

1.  Now that you have a rough idea of what the paper is about, it’s time to examine the dataset `bm`. Carry out the following steps:

    a.  Display the tibble `bm`. How many rows and columns does it have?

    b.  Display only the columns `sex`, `race` and `firstname` of `bm`. What information do these columns contain? How are `sex` and `race` encoded?

    c.  Add two new columns to `bm`: `female` should take the value `TRUE` if `sex` is female, and `black` should take value `TRUE` if `race` is black.

## Question 2a

```{r}
bm
```

4870 $\times$ 65

## Question 1b

```{r}
bm |>
  select(sex, race, firstname)

typeof(bm[["sex"]])
typeof(bm[["race"]])
typeof(bm[["firstname"]])
```

Sex (char): f - female, m - male

Race (char): w - white, b - black

firstname (char): first name of the candidate

## Question 1c

```{r}
bm <- 
  bm |>
    mutate(female = (sex == "f"),
         black = (race == "b"))
```

# Question 3

# Question 4

Randomized controlled trials are all about *balance*: when the treatment is randomly assigned, the characteristics of the treatment and control groups will be the same on average. To answer the following parts you’ll need a few additional pieces of information. First, the variable `computerskills` takes on the value `1` if a given resume says that the applicant has computer skills. Second, the variables `education` and `yearsexp` indicate level of education and years experience, while `ofjobs` indicates the number of previous jobs listed on the resume. To analyse whether characteristics are balanced across treatment and control, compare summary statistics between applicants with white- and black-sounding names.

a.  Is sex balanced across race? Use `dplyr` to answer this question. **Hint:** what happens if you apply the function `sum` to a vector of `TRUE` and `FALSE` values?

b.  Are computer skills balanced across race? **Hint:** the summary statistic you’ll want to use is the *proportion* of individuals in each group with computer skills. If you have a vector of ones and zeros, there is a very easy way to compute this.

c.  Are `education` and `ofjobs` balanced across race?

d.  Compute the mean and standard deviation of `yearsexp` by race. Comment on your findings.

e.  Why do we care if `sex`, `education`, `ofjobs`, `computerskills`, and `yearsexp` are balanced across race?

f.  Is `computerskills` balanced across `sex`? What about `education`? What’s going on here? Is it a problem? **Hint:** re-read section II C of the paper.

## Question 4a

```{r}
bm |>
  group_by(race) |>
  summarise(female_percent = sum(female) / n())
```

Almost balanced.

## Question 4b

```{r}
bm |>
  group_by(race) |>
  summarise(computer_skill_percent = sum(computerskills) / n())
```

Again, nearly balanced.

## Question 4c

```{r}
bm |>
  group_by(race) |>
  summarise(avg_eduction = mean(education),
            avg_ofJobs = mean(ofjobs))
```

Almost balanced.

## Question 4d

```{r}
bm |>
  group_by(race) |>
  summarise(mean_yearsexp = mean(yearsexp),
            sd_yearsexp = sd(yearsexp))
```

Black has a lower average years of experience with lower dispersion.

## Question 4e

## Question 4f

# Question 5

The outcome of interest in bm is call which takes on the value 1 if the corresponding resume elicits an email or telephone callback for an interview. Check your answers to the following against Table 1 of the paper:

a.  Calculate the average callback rate for all resumes in bm.

b.  Calculate the average callback rates separately for resumes with “white-sounding” and “black-sounding” names. What do your results suggest?

c.  Repeat part (b), but calculate the average rates for each combination of race and sex. What do your results suggest?

## Question 5a

```{r}
mean(bm[["call"]])
```

## Question 5b

```{r}
bm |>
  select(call, race) |>
  group_by(race) |>
  summarise(avgCallRate = mean(call))
```

## Question 5c

```{r}
bm |>
  select(call, race, sex) |>
  group_by(race, sex) |>
  summarise(avgCallRate = mean(call))
```

# Question 6

You saw in 5 (b) that applicants with African-American sounding names receive fewer callbacks. Is this statistically significant? Test the null hypothesis that there is no difference in callback rates between black- and white-sounding names against its two-sided alternative to find out.

a.  Read the help file for the dplyr function pull() and use it to extract two binary vectors of call-backs from bm, one for black names and one for white names.

b.  Store sample sizes, sample means, and sample variances in variables. Use these to compute your test statistic and degrees of freedom.

c.  Look up the function qt() and use it to compute the critical value for a two-sided t-test with $\alpha = 0.05$.

d.  Look up the function pt() and use it to compute the p-value. Check your result against Table 1.

e.  Comment on your results.

## Question 6a

# Question 7

What’s in a name? Read Section II E and Section IV of BM. Then look up “The Causes and Consequences of Distinctively Black Names” by Roland Fryer and Steven Levitt. Read its introduction and conclusion and answer these reading questions:

a.What do you think: what social content is contained in names, and what questions can we study using names?

b.  What are some weaknesses that BM acknowledge in their study?

c.  What are some potential confounds that may complicate the interpretation of results based on randomly assigning stereotypically black and white names to resumes?

d.  What is “taste-based” discrimination? What is “statistical” discrimination? How consistent are these models with the results of BM?

e.  What are the key findings of Fryer and Levitt’s study?

f.  How do Fryer and Levitt’s results relate to those of BM? What are some possible ways to reconcile the two sets of findings?
