---
title: "FREDR"
author: "Xiaotian Tian"
format: pdf
editor: visual
---

# Exercise 1

a.  Install `R`’s `fredr` package and load the library.

b.  Create a FRED account [here](https://fred.stlouisfed.org/).

c.  After creating your account, login and generate an API key. Save your key as a string in a variable called `api_key`.

d.  Set your API key using `fredr_set_key(api_key)`.

Congratulations, you can now directly retrieve macroeconomic time series from the Federal Reserve Bank!

e.  Suppose you are interested in GDP. You can get access to quarterly data from 1947 by simply tipping `fredr(series_id = "GDP")`. Try it out!

f.  We will focus on data from 2000 for this problem sheet. Type `?fredr` into the console and study the function’s documentation. Modify your API call to retrieve quarterly GDP data from January 01, 2000, and save it in a `tibble` called `gdp`. **Hint:** If you want to know what other time series `fredr` has you can use `fredr_category_children(category_id = 0)` to navigate the data directory.

g.  Next, we’ll retrieve unemployment rates. To find the correct `series_id`, use `fredr`’s text search: simply run `fredr_series_search_text("unemployment")`. After identifying the correct id, save monthly data on US unemployment rates from January 2000 in a `tibble` called `u`.

## 1a

```{r}
# install.packages("fredr")
library(fredr)
```

## 1b

```{r}
api_key <- "bc72f0649de4a3ae5a5daf4114b2aaaa" # delete this befor rendering!
fredr_set_key(api_key) # delete this befor rendering!
```
