---
title: "PS2_Wells"
format: html
editor: visual
---

# Question 1

```{r}
.rs.restartR() #restart R session
```

```{r}
options(crayon.enabled = FALSE) # suppress colorised warnings to be displayed correctly
gc() # garbage collection
rm(list = ls()) # clear variables
```

```{r}
library(tidyverse)
library(broom)
library(margins)
library(modelsummary)
```

## a

```{r}
wells <- read.csv("https://ditraglia.com/data/wells.csv")
```

## b

```{r}
wells <- wells |>
  mutate(larsenic = log(arsenic))
```

## c

```{r}
wells |>
  ggplot(aes(x = arsenic, y = larsenic)) +
  geom_point() +
  labs(
    title = "Arsenic level of respondent’s well",
    caption = "(100s of micrograms per liter))"
  ) +
  xlab("arsenic") +
  ylab("log arsenic") +
  theme_bw()
```

## d

```{r}
wells <- wells |>
  mutate(dist100 = dist / 100)
```

## e

```{r}
wells <- wells |>
  mutate(zeduc = (educ - mean(educ)) / sd(educ))
```

# Question 2

## a

```{r}
fit1 <- glm(switch ~ dist100, family = binomial(link = "logit"), wells)

summary(fit1)
```

## b

```{r}
ggplot(wells, aes(x = dist100, y = switch)) +
  stat_smooth(method = "glm", method.args = list(family = binomial)) +
  geom_jitter(width = 0.2, height = 0.1)
```

## c

`dist100` is a significant predictor of `switch` at 5% SL.

Its coefficient is negative, which makes sense because the farther from a safe well a household is, the more helpful a new well will be.

## d

```{r}
predict(fit1,
        newdata = data.frame(dist100 = mean(wells$dist100)),
        type = "response")
```

## e

```{r}
# Average Marginal Effect
summary(margins(fit1))$AME
```

```{r}
# Marginal Effect at Average
marginal_effects(fit1, data = data.frame(dist100 = mean(wells$dist100)))
```

A unit increase in `dist100` from its average is predicted to decrease the probability of building a new well by 0.1519011.

In absolute value, this is larger than the average marginal effect (-0.149842).

# Question 3

## a

```{r}
wells <- wells |>
  mutate(p1 = predict(fit1, newdata = data.frame(dist100), type = "response"))
```

## b

```{r}
wells <- wells |>
  mutate(pred1 = ifelse(p1 > 0.5, 1, 0))
```

## c

```{r}
wells |>
  mutate(success = (pred1 == switch)) |>
  pull(success) |>
  mean()
```

## d

```{r}
table(Predicted = wells$pred1, Actual = wells$switch)
```

## e

$$
\text{Sensitivity} = \frac{\text{True Positives (TP)}}{\text{True Positives (TP)} + \text{False Negatives (FN)}}
$$

```{r}
1604 / (1604 + 1089)
```

$$
\text{Specificity} = \frac{\text{True Negatives (TN)}}{\text{True Negatives (TN)} + \text{False Positives (FP)}}
$$

```{r}
194 / (194 + 1089)
```

## f

```{r}
# construct null model

# find the mode
wells |>
  group_by(switch) |>
  summarise(obs = n())
```

```{r}
# generate predictions in the null model
wells <- wells |>
  mutate(pred1_null = 1)
```

```{r}
# confusion matrix for the null model
table(Predicted = wells$pred1_null, Actual = wells$switch)
```

```{r}
list(
  accuracy = 1737 / (1283 + 1737),
  sensitivity = 1737 / (1737 + 0),
  specificity = 0 / (0 + 1283)
)
```

The null model has lower accuracy, perfect sensitivity, and zero specificity.

This illustrate the sensitivity-specificity trade-off.

# Question 4

## a

```{r}
fit2 <- glm(switch ~ larsenic, family = binomial(link = "logit"), data = wells)
```

## b

```{r}
fit3 <- glm(switch ~ zeduc, family = binomial(link = "logit"), data = wells)
```

## c

```{r}
fit4 <- glm(switch ~ dist100 + larsenic + zeduc, family = binomial(link = "logit"), data = wells)
```

## d

```{r}
all_regressions = list(fit1, fit2, fit3, fit4)

modelsummary(all_regressions,
             fmt = 2,
             stars = TRUE,
             title = "All Logit Model Estimates")
```

# Question 5

## a

```{r}
summary(fit2)
```

```{r}
ggplot(wells, aes(x = larsenic, y = switch)) +
  stat_smooth(method = "glm", method.args = list(family = binomial)) +
  geom_jitter(width = 0.2, height = 0.1)
```

`larsenic` is a significant predictor at 5% SL. The sign is positive which makes sense — as the water in the well gets more poisonous, it's more likely for a household to switch.

## b

```{r}
summary(fit3)
```

```{r}
ggplot(wells, aes(x = zeduc, y = switch)) +
  stat_smooth(method = "glm", method.args = list(family = binomial)) +
  geom_jitter(width = 0.2, height = 0.1)
```

`zeduc` is a significant predictor at 5% SL. The sign is positive which makes sense — more educated households are more likely to be aware of the water quality problem, and hence morel likely to switch.

## c

```{r}
marginal_effects(fit4,
                 data = wells |> 
                   summarise(
                     across(
                       c(dist100, larsenic, zeduc),
                       mean,
                       .names = "{.col}")
                     )
                 )
```

Those are the marginal effects on predicted switching probability at the average value of those variables.

# Question 6

## a

```{r}
# Generate p4
wells <- wells |>
  mutate(p4 = predict(
    fit4,
    newdata = data.frame(
      dist100 = dist100,
      larsenic = larsenic,
      zeduc = zeduc
    ),
    type = "response"
  ))
```

```{r}
# Generate pred4
wells <- wells |>
  mutate(pred4 = ifelse(p4 > 0.5, 1, 0))
```

```{r}
# accuracy
wells |>
  summarise(accuracy = mean(ifelse(pred4 == switch, 1, 0)))
```

```{r}
# confusion matrix
table(Predict = wells$pred4, Actual = wells$switch)
```

```{r}
# sensitivity, specificity
list(
  sensitivity = 1358 / (1358 + 379),
  specificity = 546 / (546 + 737)
)
```

## b

In our sample, `fit4` has higher accuracy, sensitivity, and specificity than `fit1`.
