---
title: "PS2_Football"
format: pdf
editor: visual
---

```{r}
options(crayon.enabled = FALSE) # suppress colorised warnings to be displayed correctly
gc() # garbage collection
rm(list = ls()) # clear variables
```

```{r}
#library packages and import dataset
library(tidyverse)
football <- read_csv('https://ditraglia.com/data/fair_football.csv')
football
```

# Question 1

## How often does the home team win?

```{r}
#generate new winning variable
football <- football |>
  mutate(team_A_wins = (SPREAD > 0))

#how often does the home team win?
football |>
  group_by(H) |>
  summarise(win_rate = mean(team_A_wins))
```

Avg win rate of the home team = 0.812

Avg win rate of the guest team = 0.502

## How many more points does the home team score on average?

```{r}
# generate home dummy
football <- football |>
  mutate(at_home = (H == 1))

# calculate means for 2 groups
q1 <- football |>
  group_by(at_home) |>
  summarise(mean_SPREAD = mean(SPREAD)) |>
  pull(mean_SPREAD)

# final result
q1[2] - q1[1]
```

# Question 2

The interpretation for intercept here is "the expected spread when both teams are in a foreign ground". However, by definition, this expectation should be zero, since team A and team B are randomly assigned.

# Question 3

```{r}
library(broom)
```

```{r}
#regression
q3_regression <- lm(SPREAD ~ factor(H) - 1, football)
summary(q3_regression)
```

Interpretation of coefficient:

-   A guest team at its opponent's home is estimated to lose 0.923 points on average. (Insignificant at 5% SL)

-   Being in a foreign field which is also unfamiliar to the opponent is estimated to win by 2.5 points on average. (Insignificant at 5% SL)

-   Being in the home field is estimated to win by 14.9518 points on average. (Significant at 5% SL)

R-squared is 0.1329.

# Question 4

```{r}
library(GGally)
```

```{r}
ggpairs(data = football,
        columns = c("MAT", "SAG", "BIL", "COL", "MAS", "DUN", "REC"))
```

All of our predictors are highly correlated.

# Question 5

```{r}
q5_regression <- lm(SPREAD ~ factor(H) + LV + REC + MAT + SAG + BIL + COL + MAS + DUN - 1, data = football)

summary(q5_regression)
```

Apart from H and LV, no other predictor has significant marginal predictive power.

Perform a joint test:

```{r}
library(car)
linearHypothesis(q5_regression, c("REC = 1", "MAT = 1", "SAG = 1", "BIL = 1", "COL = 1", "MAS = 1", "DUN = 1"))
```

They do have significant joint predictive power, so we cannot drop them.

Compare with SLR:

```{r}
SLRs <- list(
  lm(SPREAD ~ MAT - 1, data = football),
  lm(SPREAD ~ SAG - 1, data = football),
  lm(SPREAD ~ BIL - 1, data = football),
  lm(SPREAD ~ COL - 1, data = football),
  lm(SPREAD ~ MAS - 1, data = football),
  lm(SPREAD ~ DUN - 1, data = football),
  lm(SPREAD ~ LV - 1, data = football)
)

for (regression in SLRs) {
  print(summary(regression)$adj.r.squared)
}
```

The highest adjusted R-square for individual predictors is 0.4559, which is still lower than the adjusted R-square for the big model 0.4608. Therefore, putting them together increases our predictive power.

# Question 6

```{r}
q6_regression <- lm(SPREAD ~ factor(H) + LV + REC - 1, data = football)

summary(q6_regression)
```

```{r}
linearHypothesis(q6_regression, c("REC = 0", "factor(H)-1 = 0", "factor(H)0 = 0", "factor(H)1 = 0"))
```

H and REC do jointly provide additional predictive information at 5% SL.

# Question 7

```{r}
q7_regression <- lm(SPREAD ~ LV - 1, data = football)

summary(q7_regression)
```

```{r}
linearHypothesis(q7_regression, "LV = 1")
```

Efficiency of betting market implies the coefficient of LV should be 1. We cannot reject that hypothesis at 5% SL.

The R-square is 0.4562, which indicates quite good predictive power – around half of the total variations in SPREAD can be explained using LV.

# Question 8

```{r}
library(modelsummary)
```

```{r}
all_regressions <- append(SLRs, list(q3_regression, q5_regression, q5_regression_2, q6_regression, q7_regression))

modelsummary(
  all_regressions,
  gof_omit = 'Log.Lik|R2 Adj.|AIC|BIC|F',
  fmt = 2,
  title = 'Regression results for football'
  )
```
