---
title: "PS3_MLDA"
format: html
---

```{r}
options(crayon.enabled = FALSE) # suppress colorised warnings to be displayed correctly
gc() # garbage collection
rm(list = ls()) # clear variables
```

```{r}
library(haven)
library(tidyverse)
library(broom)
```

# Question 1

```{r}
# load data
data <- read_dta("https://ditraglia.com/data/mlda.dta")
```

```{r}
# construct variables
data <- data |>
  mutate(
    # center at zero
    agecell_centered  = agecell - 21,
    # cutoff variable
    D  = if_else(agecell >= 21, 1, 0)
  )
```

Model setup

$$
\text{all}_i = \alpha + \beta_0 \text{agecell_centered}_i + \beta_1\,R_i   +\beta_2\,(R_i \times \text{agecell_centered}_i) +\epsilon_i
$$

CATE: ATE on individuals with age exactly 21.

# Question 2

```{r}
# fit RD model
rd_linear <- lm(all ~ agecell_centered + D + agecell_centered:D, data = data)

# summary
summary(rd_linear)
```

```{r}
# visualise
data |>
  ggplot(aes(agecell, all, color = factor(D))) +
  geom_point() +
  geom_smooth(method = 'lm', formula = y ~ x) +
  theme(legend.position = 'none')
```

Interpretation and Inference:

-   Drinking, on average, leads to 7.66 more deaths per 100,000 individuals for individuals at age 21. This is statistically significant at 1% SL.

-   The difference slopes is not statistically significant across the cutoff, making the RD design more credible.

# Question 3

```{r}
# fit quadratic model
rd_quad_full <- lm(
  all ~ (agecell_centered + I(agecell_centered ^ 2)) * D,
  data = data
)

# summary
summary(rd_quad_full)
```

```{r}
# visualise
data |>
  ggplot(aes(agecell, all, color = factor(D))) +
  geom_point() +
  geom_smooth(method = 'lm', formula = y ~ x ^ 2) +
  theme(legend.position = 'none')
```

The result is still significant at 1% SL. Now the estimated CATE is 9.5478 more deaths per 100,000 individuals for individuals at age 21.

# Question 4

```{r}
# filter
data_filtered <- data |>
  filter(agecell >= 20, agecell <= 22)
```

```{r}
# fit RD model
rd_linear <- lm(all ~ agecell_centered + D + agecell_centered:D, data = data_filtered)

# summary
summary(rd_linear)
```

```{r}
# visualise
data_filtered |>
  ggplot(aes(agecell, all, color = factor(D))) +
  geom_point() +
  geom_smooth(method = 'lm', formula = y ~ x) +
  theme(legend.position = 'none')
```

```{r}
# fit quadratic model
rd_quad_full <- lm(
  all ~ (agecell_centered + I(agecell_centered ^ 2)) * D,
  data = data_filtered
)

# summary
summary(rd_quad_full)
```

```{r}
# visualise
data_filtered |>
  ggplot(aes(agecell, all, color = factor(D))) +
  geom_point() +
  geom_smooth(method = 'lm', formula = y ~ x ^ 2) +
  theme(legend.position = 'none')
```

Now, the corresponding CATE become 9.753 and 9.611 correspondingly. They are still significant at 5% SL. The new estimates are similar to previous ones, adding extra credibility to our result.
